\pkg{hesim}: Health Economic Simulation Modeling and Decision Analysis

### 2020-12-16 manuscript.Rnw
Some text updates

\pkg{hesim} is an \proglang{R} package that advances the functionality and performance of the existing software. Multiple model types are supported including cohort discrete time state transition models (DTSTMs), N-state PSMs, and individual-level continuous time state transition models (CTSTMs), encompassing both Markov (time-homogeneous and time-inhomogeneous) and semi-Markov processes. To maximize flexibility and facilitate integration of the statistical methods and economic model, parameters can be estimated either by fitting a model in \proglang{R} or by inputting parameters obtained from external sources. So that individual-level simulation and PSA can be run quickly, \pkg{Rcpp} and \pkg{data.table} are heavily utilized. After simulating costs and quality-adjusted life-years (QALYs) from a PSA, decision analysis can be performed within a cost-effectiveness framework.

The remainder of this article is organized as follows. Section~\ref{sec:model-taxonomy} describes the economic models supported by \pkg{hesim}. An overview of the coding framework is provided in Section~\ref{sec:framework}. The next two sections contain example analyses. Section~\ref{sec:modeling-patient} shows how both partitioned survival and multi-state models can be used in an ideal case where rich patient level data is available. Section~\ref{sec:modeling-aggregate} then moves to the common scenario where only aggregate level data is available shows how both cohort and individual models can be employed. The cost-effectiveness framework is described in Section~\ref{sec:cea} along with worked examples. Section~\ref{sec:discussion} makes comparisons to other software and discusses possible extensions. Finally, Section~\ref{sec:conclusion} concludes.

@@

Markov and semi-Markov models can be formulated in either continuous or discrete time, at either the cohort or individual level. In health economics, Markov cohort models tend to be formulated in discrete time (i.e., as cDTSTMs), although state probabilities can be computed in continuous time models using the Aalen-Johansen estimator \citep{aalen1978empirical} or the Kolmogorov forward equation \citep{cox1977theory}. While tunnel states can be used to approximate a semi-Markov model using a cohort approach, they can only be simulated in a general fashion using individual level models. Discrete time individual simulation is possible, but we use a continuous time models (i.e., iCTSTMs) because they do not require specification of model cycles and can be run considerably faster.

PSMs are specialized models that can be parameterized using survival curves and are especially useful in oncology where PFS and OS are commonly reported. They are ``area under the curve'' models, although they can also be formulate as STMs by using the survival curves to construct transition probabilities.

@@

%% -- Framework ----------------------------------------------------------------
\section{Framework} \label{sec:framework}
Economic models consist of a disease model, a utility model, and a set of cost models for each cost category. Model development proceed in 4 steps. The first step is to setup the model by defining the model structure, target population, and treatment strategies of interest, and storing the relevant information in a \code{hesim_data} object.

@@

\subsection{Parameterization} \label{sec:parameterization}
Each submodel contains fields for the model parameters and the input data. Parameters can either be created from a statistical model fit using \proglang{R} or input directly by the user. There are two types of parameter objects, standard parameter objects prefixed by \code{``params''} and transformed parameter objects prefixed by \code{``tparams''}. The former contain the underlying parameters of a statistical model while the latter contain parameters more immediate to prediction that have been transformed as function of the input data. The coefficients from a Weibull regression model are an example of a parameter object and the predicted shape and scale parameters of the Weibull distribution are an example of a transformed parameter object.

Although economic models are ideally parameterized using explicit statistical models, it is not always feasible. Parameters are often taken from a variety of sources or from summary level data. In these cases, it can be convenient to build models with mathematical expressions usin non-standard evaluation. This can be done when building cDTSTMs using \code{defime_model()}.

@@

\begin{table} [h]
\caption{Parameterization of disease models}\label{tbl:parameterize-disease-model}
\footnotesize
\begin{tabular*}{\textwidth}{l l l l}
\hline
Economic model & Statistical model & Parameter object & Model object\\
\hline
\multirow{2}{*}{\code{CohortDtstm}}& Custom & \code{tparams_transprobs} & \code{define_model()}\\
 & Multinomial logistic regressions & \code{params_mlogit} & \code{multinom_list}\\
\multirow{2}{*}{\code{IndivCtstm}} & Multi-state model (joint likelihood) & \code{params_surv} & \code{flexsurv::flexsurvreg}\\
& Multi-state model (transition-specific) & \code{params_surv_list} & \code{flexsurvreg_list}\\
\code{Psm} & Independent survival models & \code{param_surv_list} & \code{flexsurvreg_list}\\
\hline
\end{tabular*}
\end{table}

@@

Although disease models are ideally parameterized by fitting a single statistical model, this is often infeasible. Parameters may be taken from multiple sources, frequently using summary level data. In these cases it can be more convenient to define transition probability matrices using mathematical expressions relating parameters and input data to transformed parameters. Transition probabilities for cDTSTMs can be constructed in this manner using \code{define_model()}.

\subsubsection{Costs and utility} 
State values (i.e., costs and utilities) do not depend on the choice of disease model. They can currently either be modeled using a linear model or using predicted means. The latter is an example of a transformed parameter object since the predicted means are parameters that are presumably a function of the underlying parameters of a statistical model and input data.

\begin{table} [h]
\caption{Parameterization of state value models}\label{tbl:parameterize-disease-model}
\footnotesize
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}l l l l}
\hline
Statistical model & Parameter object & Model object\\
\hline
\multirow{2}{*}{Predicted means} & \code{tparams_mean} & \code{stateval_tbl}\\
 & \code{tparams_mean} & \code{define_model()}\\
Linear model & \code{params_lm} & \code{stats::lm}\\
\hline
\end{tabular*}
\end{table}

The most straightforward way to construct state values is with a \code{stateval_tbl}, which is a special object used to assign values to health states that can vary across PSA samples, treatment strategies, patients, and/or time intervals. State values can be specified either as moments (i.e., mean and standard error) or parameters (e.g., shape and scale of gamma distribution) of a probability distribution, or by pre-simulating values from a suitable probability distribution (e.g., from a Bayesian model). Like transition probabilities, state value can also be constructed with \code{define_model}.

@@

%% -- Illustrative example -----------------------------------------------------
\section{Illustrative example} \label{sec:illustrative-example}

@@

@@

@@

states <- data.table(
  state_id = c(1, 2),
  state_name = c("Stable", "Progression") # Non-death health states
)
n_states <- nrow(states)

strategies <- data.frame(
  strategy_id = 1:3,
  strategy_name = c("SOC", "New 1", "New 2"),
  soc = c(1, 0, 0),
  new1 = c(0, 1, 0),
  new2 = c(0, 0, 1)
)
n_strategies <- nrow(strategies)

@@

%% -- Individual continuous time state transition model ------------------------
\subsection{Individual continuous time state transition model}

@@

<<ictstm_fits>>=
library("flexsurv")
n_trans <- max(tmat, na.rm = TRUE) 
wei_fits <- vector(length = n_trans, mode = "list")
f <- as.formula(Surv(time, status) ~ factor(strategy_name) + female + age)
for (i in 1:length(wei_fits)){
  if (i == 3) f <- update.formula(f, .~.-factor(strategy_name)) 
  wei_fits[[i]] <- flexsurvreg(f, data = onc3, 
                               subset = (transition_id == i),
                               dist = "weibull")
}
wei_fits <- flexsurvreg_list(wei_fits)
@

@@

<<ictstm_drugcost_setup, echo=FALSE, results="hide">>=
n_times <- 2
n_i <- n_states * n_times
drugcost_dt <-  data.table(
    strategy_id = rep(strategies$strategy_id, each = n_i),
    state_id = rep(rep(states$state_id, each = n_times), n_strategies),
    time_start = rep(c(0, 3/12), n_states * n_strategies),
    est = c(rep(2000, n_i -1), 1500, 
            rep(12000, n_i -1), 1000,
            rep(15000, n_i - 1), 1200
            ) 
)
@

@@

\subsection{Cohort disctete time state transition model}

>>>>

\pkg{hesim} is an \proglang{R} package that advances the functionality and performance of the existing software. Multiple model types are supported including cohort discrete time state transition models (DTSTMs), N-state PSMs, and individual-level continuous time state transition models (CTSTMs), encompassing both Markov (time-homogeneous and time-inhomogeneous) and semi-Markov processes. To maximize flexibility and facilitate integration of the statistical methods and economic model, parameters can be estimated either by fitting a model in \proglang{R}, by inputting parameters obtained from external sources. So that individual-level simulation and PSA can be run quickly, \pkg{Rcpp} and \pkg{data.table} are heavily utilized. After simulating costs and quality-adjusted life-years (QALYs) from a PSA, decision analysis can be performed within a cost-effectiveness framework.

The remainder of this article is organized as follows. Section~\ref{sec:model-taxonomy} describes the economic models supported by \pkg{hesim}. An overview of the coding framework is provided in Section~\ref{sec:framework}. An illustrative example using a three state model of disease progression in oncology is provided in Section~\ref{sec:illustrative-example}. Analyses are conducted for each of the three supported model types, illustrating approaches suitable for both patient level and summary level data. The cost-effectiveness framework is described in Section~\ref{sec:cea} along with worked examples based on the output from the prior section. Section~\ref{sec:discussion} makes comparisons to other software and discusses possible extensions. Finally, Section~\ref{sec:conclusion} concludes.

@@

Markov and semi-Markov models can be formulated in either continuous or discrete time, at either the cohort or individual level. In health economics, Markov cohort models tend to be formulated in discrete time (i.e., as cDTSTMs), although state probabilities can be computed in continuous time models using the Aalen-Johansen estimator \citep{aalen1978empirical} or the Kolmogorov forward equation \citep{cox1977theory}. While tunnel states can be used to approximate a semi-Markov model using a cohort approach, they can only be simulated in a general fashion using individual level models. Discrete time individual simulation is possible, but we use continuous time models (i.e., iCTSTMs) because they do not require specification of model cycles and can be run considerably faster.

PSMs are specialized models that can be parameterized using survival curves and are especially useful in oncology where PFS and OS are commonly reported. They are ``area under the curve'' models, although they can also be formulated as STMs by using the survival curves to construct transition probabilities.

@@

%% -- Framework ----------------------------------------------------------------
\section{Framework} \label{sec:framework}
Economic models consist of a disease model, a utility model, and a set of cost models for each cost category. Model development proceeds in 4 steps. The first step is to setup the model by defining the model structure, target population, and treatment strategies of interest, and storing the relevant information in a \code{hesim_data} object.

@@

\subsection{Parameterization} \label{sec:parameterization}
Each submodel contains fields for the model parameters and the input data. Parameters can either be created from a statistical model fit using \proglang{R} or input directly by the user. There are two types of parameter objects, standard parameter objects prefixed by \code{``params''} and transformed parameter objects prefixed by \code{``tparams''}. The former contain the underlying parameters of a statistical model while the latter contain parameters more immediate to prediction that have been transformed as function of the input data. The regression coefficients of a logistic regression are example of a parameter objects while the predicted probabilities are examples of a transformed parameter object.

Transformed parameter objects are often useful when parameterizing a model using external sources. Two common examples are predicted probabilities for transition matrices and predicted means for assigning values to health states. The advantage of a transformed parameter object is that an explicit statistical model does not need to be specified which affords more flexibility to the user.

@@

\begin{table} [h]
\caption{Parameterization of disease models}\label{tbl:parameterize-disease-model}
\footnotesize
\begin{tabular*}{\textwidth}{l l l l}
\hline
Economic model & Statistical model & Parameter object & Model object\\
\hline
\code{CohortDtstm} & Custom & \code{tparams_transprobs} & \code{msm::msm} \\
 & Multinomial logistic regressions & \code{params_mlogit} & \code{multinom_list}\\
\code{IndivCtstm} & Multi-state model (joint likelihood) & \code{params_surv} & \code{flexsurv::flexsurvreg}\\
& Multi-state model (transition-specific) & \code{params_surv_list} & \code{flexsurvreg_list}\\
\code{Psm} & Independent survival models & \code{param_surv_list} & \code{flexsurvreg_list}\\
\hline
\end{tabular*}
\end{table}

@@

\code{tparams_transprobs} objects contain transition probability matrices that have been predicted for each PSA sample, treatment strategy, and patient from the target population. Transition probabilities may also vary by time interval to allow for time inhomogeneous Markov models. Transition probabilities can be predicted from a fitted multi-state model using the \pkg{msm} package or constructed ``by hand'' in a custom manner.

\subsubsection{Costs and utility} 
State values (i.e., costs and utilities) do not depend on the choice of disease model. They can currently either be modeled using a linear model or using predicted means.

\begin{table} [h]
\caption{Parameterization of state value models}\label{tbl:parameterize-disease-model}
\footnotesize
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}l l l l}
\hline
Statistical model & Parameter object & Model object\\
\hline
Predicted means & \code{tparams_mean} & \code{stateval_tbl::lm}\\
Linear model & \code{params_lm} & \code{stats::lm}\\
\hline
\end{tabular*}
\end{table}

The most straightforward way to construct state values is with a \code{stateval_tbl}, which is a special object used to assign values (i.e. predicted means) to health states that can vary across PSA samples, treatment strategies, patients, and/or time intervals. State values can be specified either as moments (i.e., mean and standard error) or parameters (e.g., shape and scale of gamma distribution) of a probability distribution, or by pre-simulating values from a suitable probability distribution (e.g., from a Bayesian model).

@@

%% -- Illustrative example -----------------------------------------------------
\section{Illustrative example} \label{sec:illustrative-example}
We consider a three state model commonly used describe disease progression in oncology. As shown in Figure~\ref{fig:three-state-model}, the three health states are stable disease (i.e. not progressed), progressed, and death. Patients can transition to a more severe health state (stable $\rightarrow$ progressed, stable $\rightarrow$ death, and progression $\rightarrow$ death) but cannot recover to a less severe health state.

@@

This model can be simulated using either of the model types described in Section~\ref{sec:model-taxonomy}. We demonstrate use of all three below, starting with the iCTSTM.

%% -- Individual continuous time state transition model ------------------------
\subsection{Individual continuous time state transition model}
We setup the model by defining the model structure, target population, and treatment strategies of interest. The transitions are characterized by a matrix where each element denotes a transition from a row to a column. If a transition is not possible it is marked with \code{NA}; otherwise, an integer denotes the transition number.

@@

The target population, treatment strategies, and (non-death) health states are stored in a \code{hesim_data} object. There are 3 treatment strategies: standard of care (SOC) and two new interventions (New 1 and New 2). The target population consists of a heterogeneous population of 1000 patients, which is large enough to ensure that averages across patients in the individual simulation are reasonably stable.

@@

states <- data.table(
  state_id = c(1, 2),
  state_name = c("Stable", "Progression") # Non-death health states
)

strategies <- data.frame(
  strategy_id = 1:3,
  strategy_name = c("SOC", "New 1", "New 2"),
  soc = c(1, 0, 0),
  new1 = c(0, 1, 0),
  new2 = c(0, 0, 1)
)

@@

\subsubsection{Parameterization}
In a STM, the disease model governs transitions between health states. For now, we assume that that patient level data with exact transition times is available so that a multi-state statistical model can be used to parameterize the disease model.

Multi-state data consists of one row for each possible transition from a given health state where only one transition is observed and all others are censored. We use the simulated dataset \code{onc3} from the \pkg{hesim} package.

@@

Multi-state models can be fit in one of two ways. First, a joint survival model with interaction terms for different transition can be estimated, which is useful if there are constraints in the parameters across
transitions, such as coefficients that are assumed equal across transitions. Here, we will take a second approach that is computationally more efficient and fit a separate model for each transition. We illustrate with parametric Weibull models but multiple distributions should be compared in a real application \citep{williams2017cost}. The treatment effect is assumed to only influence transitions from stable disease since patients are assumed to switch to a second line treatment after progression.

<<ictstm_fits, warning = FALSE>>=
library("flexsurv")
n_trans <- max(tmat, na.rm = TRUE) 
wei_fits <- vector(length = n_trans, mode = "list")
f <- as.formula(Surv(time, status) ~ factor(strategy_name) + female + age)
for (i in 1:length(wei_fits)){
  if (i == 3) f <- update.formula(f, .~.-factor(strategy_name)) 
  wei_fits[[i]] <- flexsurvreg(f, data = onc3, 
                               subset = (transition_id == i),
                               dist = "weibull")
}
wei_fits <- flexsurvreg_list(wei_fits)
@

@@

<<ictstm_drugcost_setup, echo=FALSE, results="hide">>=
n_times <- 2
n_states <- nrow(states)
n_strategies <- nrow(strategies)
n_i <- n_states * n_times
drugcost_dt <-  data.table(
    strategy_id = rep(strategies$strategy_id, each = n_i),
    state_id = rep(rep(states$state_id, each = n_times), n_strategies),
    time_start = rep(c(0, 3/12), n_states * n_strategies),
    est = c(rep(2000, n_i -1), 1500, 
            rep(12000, n_i -1), 1000,
            rep(15000, n_i - 1), 1200
            ) 
)
@

@@

\subsection{Cohort disctete time state transition model} \label{sec:cdtstm-example}
